MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-20 results.tex
Group revisions

\section{Results}
In the Top8000 system there are a total of 235 named rotamers which reach above the 0.3\% outlier cutoff and include $\ge$ 4 datapoints. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see section \ref{sec_md:chi_stats}), which improves on both the modal and the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: the table in Section S4 describes the frequency and count of each rotamer in each residue type and the tables in Section S5 have central dihedral and covalent angle values for each rotamer. The latter numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in section \ref{sec_md:contours}.

The order-of-magnitude increase in size and quality of our reference dataset greatly improves signal-to-noise, allowing us to do a better job of the same functions as before. More importantly, it enables new features and new conclusions. The simplest view of a rotamer is of a favored sidechain conformation described by the list of dihedral angles, with ideal staggered $\chi$ values between sp$^{3}$ atoms and little preference for $\chi$ between sp$^{3}$ and sp$^{2}$ atoms. The next level of detail adds a standard deviation for each $\chi$. However, rotamers are really much more complex, especially in a macromolecular interior where numerous interactions can compensate for the energy required to depart from ideal. We have high confidence in the reliability of our more complex and nuanced conformational distributions, as we took great care to include only physically plausible sidechains that fit clear electron density (see Section \ref{sec:resfilters}). As Figure \ref{fig:outcontour} demonstrates, sidechain conformations do cluster in discrete regions of $\chi$ space. Further, there is large variance in $\chi$ values within a rotamer cluster, the shape is usually not axis-oriented, and its center is rarely at the nominally ideal staggered conformation. A rotamer is better thought of as a local energy well, with a potentially complex shape, describing the favorable extent around an allowed sidechain conformation.

@@

For validation purposes, users are often most interested in the binary issue of whether a sidechain conformation is rotameric or an outlier, i.e. falls within or outside the rotamer distribution. With the Top500 data, contours were smooth and reproducible (between different datapoint selections) only out to the 1\% level. Due to greater numbers and increased reliability in the filtered Top8000 distributions, a lower outlier cutoff is now feasible as well as desirable, since an even lower percentage of the reference datapoints are now dubious. We also preferred to keep outlier numbers for unfiltered, general data roughly equal between the previous and new systems. Therefore a count of the 1\% outliers by residue type in the unfiltered dataset was performed using the Top500 contours, and those numbers were compared to outlier counts at several different Top8000 contour cutoffs. A new cutoff of 0.3\% matched best, was found to behave smoothly, and happens also to match the 3$\sigma$ level for a normal distribution (3 out of 1000). \textcolor{changecolor}{Table S2 reports the outlier counts in the unfiltered dataset for both the Top500 and Top8000 contours.} Further, along with favored and outlier, a new category has now been added: \textit{allowed}, for scores $\ge$0.3\% and $<$2.0\%. This new category lets users know if a given conformation is at the edge of the given rotamer's distribution, close to the outlier region. This change matches the 3-part system, and the division at 2.0\%, long judged useful for Ramachandran criteria \cite{Kleywegt1996, Hooft1996, Lovell:2003uq}.

@@

\subsection{Dihedral and Bond Angle Deviations}
As we have shown, rotamers are more complex than just a collection of mean $\chi$ values and standard deviations. The allowed regions are often large and complex, and the mean or modal $\chi$ angles can \textcolor{changecolor}{deviate significantly from the expected staggered conformation}. Such deviations usually occur in well-packed environments where the sidechain makes close atomic interactions, either van der Waals (vdW) or a hydrogen bond, with its own backbone (or within the sidechain). The $\chi$ deviations help allow the contact to be close rather than a steric clash.

@@

It has long been known that certain sidechain conformations can cause large deviations in the otherwise tight covalent bond angles \cite{lovell2000penultimate}; this is seen in the scenarios above where sidechain-backbone interactions occur. One rotamer of this type is methionine \textbf{ppp}, which has just 48 examples (0.29\% of Met) in the filtered Top8000. Figure \ref{fig:2bmo} shows one specific example and demonstrates how deviations in \textit{both} $\chi$ dihedrals and covalent bond angles are needed to avoid a clash and form a favorable non-bonded interaction. All-atom contacts demonstrate the tight vdW packing. For this residue, the angles C-C$\alpha$-C$\beta$, C$\alpha$-C$\beta$-C$\gamma$, and C$\beta$-C$\gamma$-S$\delta$ each open up from Engh \& Huber values, by +1.6$\degree$, +0.9$\degree$, and +3.0$\degree$ respectively.

@@

For the aspartate and asparagine contours shown in Figure \ref{fig:AspAsn}, sometimes $\chi_{2}$ has disallowed regions and sometimes the entire range is allowed. This fact makes it even more difficult, and less meaningful, to assign a central value and standard deviation that adequately describes the allowed conformations. Asp and Asn also show clearly that there is distinct, complex fine-structure of local clustering in the data distributions, especially evident for Asn and for the diagonally streaked shoulder at the top of Asp $\chi_{1}$ \textbf{t}. These datapoint clusters are primarily due to patterns of sidechain H-bonding to specific backbone donors or acceptors local in sequence, or occasionally due to especially favorable vdW packing against local backbone.  For example, the Asn N$\delta$2 can either H-bond with the i-4 CO in a regular $\alpha$-helix in an \textbf{m-80} conformer or pack against that CO in the more common \textbf{m-20} conformer \cite{Lovell1999}.

>>>>

\section{Results}
In the Top8000 system there are a total of 235 named rotamers which reach above the 0.3\% outlier cutoff and include $\ge$ 4 datapoints. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see Section \ref{sec_md:chi_stats}), which we feel represents the clusters better than either the modal or the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: the table in Section S4 describes the frequency and count of each rotamer in each residue type and the tables in Section S5 give central dihedral and covalent angle values for each rotamer. The latter numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in Section \ref{sec_md:contours}.

The order-of-magnitude increase in size and quality of our reference dataset greatly improves signal-to-noise, allowing us to do a better job of the same functions as before. More importantly, it enables new features and new conclusions. The simplest view of a rotamer is of a favored sidechain conformation described by the list of dihedral angles, with ideal staggered $\chi$ values between sp$^{3}$ atoms and little preference for $\chi$ between sp$^{3}$ and sp$^{2}$ atoms. The next level of detail adds a standard deviation for each $\chi$. However, conformational preferences are really much more complex, especially in a macromolecular interior where numerous interactions can compensate for the energy required to depart from ideal. We have high confidence in the reliability of our more nuanced conformational distributions, as we took great care to include only physically plausible sidechains that fit clear electron density (see Section \ref{sec:resfilters}). As Figure \ref{fig:outcontour} demonstrates, sidechain conformations do cluster in discrete regions of $\chi$ space. Further, there is large variance in $\chi$ values within a rotamer cluster, the shape is usually not axis-oriented, and its center is rarely at the nominally ideal staggered conformation. A rotamer is better thought of as a local energy well, with a potentially complex shape, describing the favorable extent around an allowed sidechain conformation.

@@

For validation purposes, users are often most interested in the binary issue of whether a sidechain conformation is rotameric or an outlier, i.e. falls within or outside the rotamer distribution. With the Top500 data, contours were smooth and reproducible (between different datapoint selections) only out to the 1\% level. Due to greater numbers and increased reliability in the filtered Top8000 distributions, a lower outlier cutoff is now feasible as well as desirable, since an even lower percentage of the reference datapoints are now dubious. We also preferred to keep outlier numbers for unfiltered, general data roughly equal between the previous and new systems. Therefore a count of the 1\% outliers by residue type in the unfiltered dataset was performed using the Top500 contours, and those numbers were compared to outlier counts at several different Top8000 contour cutoffs. A new cutoff of 0.3\% matched best, was found to behave smoothly, and happens also to match the 3$\sigma$ level for a normal distribution (3 out of 1000). \textcolor{changecolor}{The table in Section S3 reports the outlier counts in the unfiltered dataset for both the Top500 and Top8000 contours.} Further, along with favored and outlier, a new category has now been added: \textit{allowed}, for scores $\ge$0.3\% and $<$2.0\%. This new category lets users know if a given conformation is at the edge of the given rotamer's distribution, close to the outlier region. This change matches the 3-part system, and the division at 2.0\%, long judged useful for Ramachandran criteria \cite{Kleywegt1996, Hooft1996, Lovell:2003uq}.

@@

\subsection{Dihedral and Bond Angle Deviations}
As we have shown, rotamers are more complex than just a collection of mean $\chi$ values and standard deviations. The allowed regions are often large and complex, and the mean or modal $\chi$ angles can \textcolor{changecolor}{deviate substantially from the expected staggered conformation}. Such deviations usually occur in well-packed environments where the sidechain makes \textcolor{changecolor}{tight atom-atom} interactions, either via van der Waals (vdW) or H-bonding, with its own backbone (or within the sidechain). The $\chi$ deviations help allow the contact to be \textcolor{changecolor}{a small overlap} rather than a steric clash.

@@

It has long been known that certain sidechain conformations show large deviations in the otherwise narrowly distributed covalent bond angles \cite{lovell2000penultimate}; this is seen in the scenarios above where sidechain-backbone interactions occur. One rotamer of this type is methionine \textbf{ppp}, which has just 48 examples (0.29\% of Met) in the filtered Top8000. Figure \ref{fig:2bmo} shows one specific example and demonstrates how deviations in \textit{both} $\chi$ dihedrals and covalent bond angles are needed to avoid a clash and form a favorable non-bonded interaction. All-atom contacts demonstrate the tight vdW packing. For this residue, the angles C-C$\alpha$-C$\beta$, C$\alpha$-C$\beta$-C$\gamma$, and C$\beta$-C$\gamma$-S$\delta$ each open up from Engh \& Huber values, by +1.6$\degree$, +0.9$\degree$, and +3.0$\degree$ respectively.

@@

For the aspartate and asparagine contours shown in Figure \ref{fig:AspAsn}, sometimes $\chi_{2}$ has disallowed regions and sometimes the entire range is allowed. This fact makes it even more difficult, and less meaningful, to assign a central value and standard deviation that adequately describes the allowed conformations. Asp and Asn also show clearly that there is distinct, complex fine-structure of local clustering in the data distributions; \textcolor{changecolor}{note especially the smaller elongated clusters at top right of $\chi_{1}$ \textbf{t}}. These datapoint clusters are primarily due to patterns of sidechain H-bonding to specific backbone donors or acceptors local in sequence, or occasionally due to especially favorable vdW packing against local backbone.  For example, the Asn N$\delta$2 can either H-bond with the i-4 CO in a regular $\alpha$-helix in an \textbf{m-80} conformer or pack against that CO in the more common \textbf{m-20} conformer \cite{Lovell1999}.
